---
title: "Trust Dimensionality Through the Lens of Polarisation: A Three-Country Study"
format: 
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    pdf-engine: xelatex
    include-in-header: preamble.tex
    geometry: margin=1in

author:
  - name: Santiago Ventura
    email: Santiago.ventura@warwick.ac.uk
    orcid: 0009-0000-4891-6091
    affiliation: 
      - id: warwick
        name: Behavioural Science Group, Warwick Business School, University of Warwick
        city: Coventry
        country: United Kingdom
    corresponding: true
  - name: Tim Mullett
    orcid: 0000-0002-4082-2813
    affiliation:
      - ref: warwick
  - name: Nick Chater
    orcid: 0000-0002-9745-0686
    affiliation:
      - ref: warwick
  - name: Cecilie Steenbuch Traberg
    orcid: 0000-0002-5640-9273
    affiliation: 
      - id: cbs
        name: Copenhagen Business School
        city: Copenhagen
        country: Denmark
abstract: |
  This paper investigates the dimensionality of citizen trust in politicians, specifically examining whether it is perceived as a three-dimensional construct (competence, benevolence, integrity) or reduces to a single, undifferentiated dimension, as a result of polarisation. Drawing on dual-process theories of cognition, we argue that the depth of cognitive processing influences trust dimensionality. We propose that in politically polarised environments, which often promote heuristic (System 1) thinking, individuals are less likely to engage in the deliberative (System 2) processing required to discriminate between three dimensions of trust, leading to a simplified, one-dimensional perception. Conversely, in less polarised contexts, where System 2 engagement is more prevalent, we anticipate the three dimensions of trust to emerge. We present findings from an initial study in the United States, in which we show evidence for dimensionality reduction. However, we find that individuals can still adjust their judgments when exposed to new information. We then present two cross-cultural replication studies, one in Chile, a similarly polarised country where we expected similar dimensionality reduction, and Denmark, a less polarised country where we anticipated finding three distinct dimensions. Contrary to our expectations, all three countries yielded a unidimensional trust structure. This research contributes to the understanding of trust formation, the cognitive effects of political polarisation, and the cross-cultural generalizability of trust models.
keywords: ["trust", "polarisation", "dimensionality"]
bibliography: "My Library.bib"

lang: en
toc: true
number-sections: true
fig-cap-location: bottom
tbl-cap-location: top

execute:
  echo: false
  warning: false
  message: false
  fig-width: 7
  fig-height: 5
---

```{r setup, include=FALSE}
library("papaja")
library("stargazer")
library("tidyr")
library("dplyr")
library("psych")
library("gridExtra")
library("readr")
library("afex")
library("emmeans")
library("lme4")
library("car")
library("patchwork")
library("sjPlot")
library("texreg")
library("stringr")
library("ggplot2")
library("Hmisc")
library("forcats")

r_refs("r-references.bib")
```

# Introduction

Trust is widely regarded as a foundational element for effective governance and social cohesion. It plays a crucial role in driving compliance and strengthening confidence in regulated sectors [@grimmelikhuijsenValidatingScaleCitizen2017; @mamanMeasuringCitizenTrust2025]. Indeed, high trust in institutions and leaders promotes citizen compliance, cooperation, and the smooth implementation of policies [@jonesConceptualFrameworkInsights2021] and is a key component of social capital that underpins democratic stability [@boixCapitalSocialDemocracia2000; @putnamMakingDemocracyWork1993].

We can understand trust as "the willingness of a party to be vulnerable to the actions of another party based on the expectation that the other will perform a particular action important to the trustor, irrespective of the ability to monitor or control that other party" [@mayerIntegrativeModelOrganizational1995].

However, recent years have seen a decline in political trust amid rising partisan polarisation. Polarisation has eroded the trust that holds democracy together, weakening faith in elected officials and even everyday interpersonal trust [@iyengarOriginsConsequencesAffective2019; @vanprooijenPsychologyPoliticalPolarization2021]. This affective animosity is rooted in partisanship as a core social identity. The source of polarisation is "not ideological but affective, based on identity" [@segoviaAffectivePolarizationLowpartisanship2022a].

In today's hyper-polarised climate, citizens often trust people only if those figures share their political identity [@leeSocialTrustPolarized2022]. This affective divide decreases people's willingness to cooperate for good causes, raising questions about how trust and partisanship jointly influence democratic decision-making. This context raises urgent questions about how people form trust judgments, and whether those judgments are as deliberate as classical theories assume.

The challenge for researchers lies in understanding how these foundational, often implicit, trust mechanisms evolve or are overridden by more explicit, deliberative judgments in complex social and political environments, particularly under increasing political polarisation. This interplay highlights why understanding the dimensionality of trust under polarisation is not merely an academic exercise but is crucial for addressing real-world challenges in governance and social cohesion.

Despite its acknowledged importance, a significant gap persists in the systematic measurement of trust [@glaeserMeasuringTrust2000]. This persistent difficulty in systematically measuring trust is not merely a methodological problem. It reflects the inherent cognitive complexity of trust itself. If trust judgments are often intuitive and non-deliberative, as suggested by the observation that individuals sometimes do not think deeply about whether they trust someone [@nisbettTellingMoreWe1977; @hirstCognitiveAspectsConsciousness1995], then explicit, rationalized survey measures might fail to capture the underlying mechanisms or the full dimensionality of how trust is actually experienced and acted upon. This indicates that the measurement challenge is intrinsically linked to the cognitive processes of trust formation.

## Trust Dimensionality

The three dimensions of trust, perceived competence, perceived benevolence, and perceived integrity, are well-established in the broader organizational trust literature. These dimensions are conceptualized as key elements forming the basis for trusting intentions and behavior.

-   Competence (or Ability): This dimension refers to the belief that the trustee possesses the necessary skills, abilities, and expertise to perform effectively within a specific domain.

-   Benevolence: This dimension describes the perception that the trustee genuinely cares about the trustor's welfare and intends to act in their best interest, extending beyond mere self-interest.

-   Integrity: This dimension represents the belief that the trustee adheres to a set of acceptable principles, is honest, and reliably keeps commitments.

These dimensions were notably validated in the *"Citizen Trust in Government Organizations"* scale [@grimmelikhuijsenValidatingScaleCitizen2017], which resulted in a nine-item measure for these three dimensions. This scale builds upon and adapts established organizational trust research, including works by @mcevilyMeasuringTrustOrganisational2011, @mcknightDevelopingValidatingTrust2002, and @mayerIntegrativeModelOrganizational1995, in public administration. @mamanMeasuringCitizenTrust2025 systematically applies and reviews the measurement of these dimensions within regulatory agencies, rather than originating the conceptualization of the dimensions themselves. This distinction is crucial for accurately crediting the theoretical lineage and demonstrating a deep understanding of the literature.

While the @mayerIntegrativeModelOrganizational1995 model was originally proposed for interpersonal and business relationships, its applicability to government organizations has been explicitly assessed in other research. It is also recognized that "it cannot be assumed that the conceptualization and measures of citizen trust in other public institutions are suited to the study of trust in regulatory agencies and can be used as a proxy to it" [@mamanMeasuringCitizenTrust2025]. This implies that while the three dimensions are theoretically robust, their manifestation or relative importance might differ when applied to abstract entities like government agencies or political figures, especially in varying political climates, which underscores the importance of empirical validation in specific contexts [@mamanMeasuringCitizenTrust2025].

## The Role of Cognitive Processing in Trust Judgments: A Dual-Process Approach

Regarding trust, @stoltzDeliberateTrustIntuitive2018 notes that *"any coherent conceptualization of the relevant phenomena has to take into account the dual nature of trust/confidence as implying sometimes reflexive and calculative stances versus more intuitive and 'faith-like' processes"*. This observation seems to align with dual-process theories of cognition. These theories posit two distinct systems of thought that guide human judgment and decision-making:

-   System 1: This system is characterized as fast, automatic, intuitive, unconscious, and effortless. It operates based on patterns, associations, and past experiences, enabling quick judgments [@kahnemanThinkingFastSlow2011]. In the context of trust, this can manifest as "intuitive faith."

-   System 2: In contrast, this system is slow, deliberate, conscious, rule-based, and effortful. It is engaged for complex problem-solving and analytical tasks, requiring intentional cognitive exertion [@kahnemanThinkingFastSlow2011]. This mode of thinking can manifest as "deliberate trust" [@stoltzDeliberateTrustIntuitive2018].

Intuitive faith is posited as the more ubiquitous and efficient form of reliance in daily life, often based on procedural knowledge and automatization, while deliberate trust, being resource-intensive, is empirically less common [@stoltzDeliberateTrustIntuitive2018].

## Polarisation and Its Impact on Social Cognition and Trust

Political polarisation is defined as the divergence of political attitudes away from the center, towards ideological extremes. This encompasses both ideological polarisation (differences in policy positions) and affective polarisation (an emotional dislike and distrust of political out-groups). Affective polarisation, in particular, involves increasingly positive feelings towards in-group members and negative feelings towards out-group members.

Polarisation significantly impacts individual cognition and social perception. It leads to several cognitive phenomena:

-   Motivated Reasoning: This refers to the tendency of people to conform assessments of information to some goal or end extrinsic to accuracy [@kahanIdeologyMotivatedReasoning2013; @kundaCaseMotivatedReasoning1990; @balcetisSeeWhatYou2006; @dunningNewerLookMotivated1999; @dittoMotivatedMoralReasoning2009]. Views held to satisfy psychological needs become valued possessions, leading individuals to reject contradictory evidence and preferentially access information that supports their desired worldview [@vanbaarPolarizedMindContext2022].

-   Biased Processing: This refers to how new information is interpreted in a partisan way, even without conscious motivation [@newmanClimateChangeCultural2018]. Our minds build meaning from information by integrating it with prior knowledge. This process can lead to biased perceptions even if the same information is consumed [@vanbaarPolarizedMindContext2022].

-   Affective polarisation: This refers to a growing dislike, distrust and hostility between members of opposing political groups, viewing the opposing side as a threat or enemy rather than simply a political rival, and an increased social distance between those with differing political views. This fosters hostility, dehumanization, and an "us vs. them" mindset, which in turn erodes trust in democratic institutions. This phenomenon is explained by Social Identity Theory and the Expressive Model of partisanship, which suggest that group affiliation can override rational decision-making [@iyengarOriginsConsequencesAffective2019; @iyengarAffectNotIdeology2012; @boxellCrossCountryTrendsAffective].

These characteristics of political polarisation are all hints of System 1 processing. This suggests that polarisation actively promotes reliance on fast, intuitive judgments and suppresses the effortful, deliberative processes of System 2. If System 2 is suppressed, individuals are less likely to engage in the deliberative evaluation required to differentiate the three dimensions of trust, leading to a simplified, one-dimensional perception. The strong emotional attachment to polarised content and the resistance to opposing views reinforces System 1 pervasiveness, making it difficult for System 2 to intervene and differentiate trust dimensions.

## The Impact of Polarisation on Trust in Leaders and Institutions

One well-documented consequence of polarisation is a decline in political trust [@ortiz-ospinaTrust2016], especially across party lines. Citizens tend to trust government institutions and officials only when their preferred party is in power [@leeSocialTrustPolarized2022]. Over the past decades, this partisan gap in trust has widened dramatically. For example, Americans whose party is out of power now express strikingly low trust in a government led by the opposing party [@hetheringtonWhyWashingtonWont2020]. Perceptions of rampant division themselves can further debilitate institutional trust. Recent studies find that when people perceive high polarisation, they become less trusting of political leaders and government generally [@leeSocialTrustPolarized2022]. In other words, people endorse co-partisan representatives while being skeptical of those from the opposing side. Affective polarisation also breeds interpersonal distrust beyond institutions. Partisans not only doubt opposing leaders – they are often reluctant to even engage with opposing supporters [@iyengarOriginsConsequencesAffective2019]. Heightened animosity leads individuals to assume the worst about out-group members' motives and credibility. This social dimension of mistrust can reinforce political inefficiencies: if voters broadly distrust those who think differently, it becomes difficult to build the public consensus needed to tackle shared problems. Thus, rising polarisation has weakened the trust that holds democracy together, from faith in elected officials to everyday trust between people. This polarisation undermines faith in democratic institutions and weakens governance, as people become increasingly unwilling to support compromises or policies advanced by the "other side" [@hetheringtonWhyWashingtonWont2020].

## Beyond the U.S.: Polarisation in Other Contexts

Much of the seminal work on polarisation, trust, and policy cueing comes from U.S. studies, but similar dynamics appear in other polarised contexts. Chile, for example, has historically had lower levels of formal party identification, yet it has experienced notable increases in affective polarisation in recent years [@segoviaAffectivePolarizationLowpartisanship2022a]. Survey research tracking Chilean public opinion from 1990 to 2021 finds that dislike between the left and right has persisted or even grown, despite many citizens no longer strongly identifying with political parties [@segoviaAffectivePolarizationLowpartisanship2022a]. In other words, even in a society where partisan ties have weakened, political factions still exhibit in-group favoritism and out-group animosity. This suggests that polarisation can manifest through broad ideological or cultural camps, producing effects akin to those in the U.S. two-party system [@EdelmanTrustBarometer2023; @segoviaAffectivePolarizationLowpartisanship2022a]. Consequently, the implications for trust and policy acceptance in Chile parallel those observed in the United States [@carlinPoliticalCompetitionPartisanship2018]. Polarised Chilean citizens may be less inclined to trust institutions when control lies with their ideological rivals, echoing patterns of partisan distrust seen in the U.S. [@leeSocialTrustPolarized2022]. These cross-national parallels underscore that the phenomena of political trust decision making are not unique to the U.S. context—they can emerge wherever political identities become salient and polarised.

## Present Research

The present studies examine how political polarisation shapes trust dimensionality, focusing on the U.S., Chile, and Denmark. Chile is a highly polarised Latin American context, and Denmark is a country known for its high levels of interpersonal trust and low polarisation, where we expected to find the three dimensions of trust. Drawing on the theories of motivated reasoning and social identity, we investigate partisan bias in evaluations of trust of political figures. With the following experiments, we aim to clarify how affective polarisation restructures political trust perceptions and deliberation.

While political trust is often conceptualized as multidimensional, encompassing perceptions of competence, integrity, and benevolence [@grimmelikhuijsenValidatingScaleCitizen2017], it remains unclear whether citizens actually rely on these distinctions when evaluating polarised leaders. Our first study tests whether partisan identity simplifies or collapses the dimensional structure of trust, effectively turning trust into a uni-dimensional in-group/out-group evaluation. This addresses a key gap in the trust literature, which has not yet empirically assessed how affective polarisation distorts trust dimensionality. In our first study, we test the dimensionality of trust perceptions when respondents evaluate prominent political figures (e.g., Biden and Trump) across party lines. We find evidence that polarisation reduces the complexity of trust judgments. We also find evidence that partisanship strongly determines trust in politicians.

Despite extensive research on trust and political polarisation, the direct interplay between political polarisation, cognitive processing styles, and the perceived dimensionality of trust remains underexplored. Specifically, there is a gap in understanding how the political environment influences the cognitive architecture of trust judgments.

Our paper outlines two further studies: one in Chile and one in Denmark. Chile provides a valuable replication context as a society that has faced intense political conflicts and increasing polarisation in recent years (e.g., during debates over its constitution and inequality, @ChileanSocietyRemains2023). If polarisation is indeed driving one-dimensional trust judgments, we expect to replicate the U.S. results in Chile, observing a similar dimensionality reduction of trust. Finally, we planned a comparative study in Denmark, a country often called "the world's most trusting" society with a consensus-oriented, multi-party political culture. Denmark has relatively high social trust and lower levels of partisan affective polarisation compared to the U.S. [@williamsDenmarkSecretHow2024]. In this context, we anticipated that citizens would exhibit a more deliberate judgment of trust, aligning with the theoretical three-factor model. In other words, Denmark's respondents should distinguish competence, benevolence, and integrity when evaluating leaders, resulting in a three-dimensional trust structure.

In summary, our research seeks to answer: To what extent is trust truly a three-dimensional construct in practice, and how does this depend on the polarisation of the environment? We address this question through cross-national studies that combine survey experiments and Principal Component Analyses. We hypothesized that polarisation affects trust's dimensionality: in highly polarised environments (U.S., Chile) trust would appear largely unidimensional, whereas in less polarised environments (Denmark) the three dimensions would be more discernible. By testing this, our work contributes to both the trust literature and the polarisation literature, examining whether a classic model of trust holds universally or breaks down under partisan bias.

# Study 1

## Methods

All methods were pre-registered in [aspredicted.org #168963](https://aspredicted.org/2h8sa.pdf).

## Participants

```{r s1-load, include=FALSE}
df1 <- read.csv("study1.csv") 

mean_dur_s1 <- mean(df1$seconds)
sd_dur_s1 <- sd(df1$seconds)
mean_age_s1 <- mean(df1$age)
sd_age_s1 <- sd(df1$age)
```

We recruited 200 participants from *Prolific* between March 21, 2024 and April 2, 2024. In particular, we aimed for a balanced sample of 100 Republican and 100 Democrat participants from the United States, so we set these as explicit inclusion/recruitment criteria for the sample, based upon participant's self-reported affiliation in their *Prolific* profile. Participants were paid £0.75 for participating in the study which was aimed to last 5 minutes. The actual duration was `r round(mean_dur_s1, 2)` seconds (SD = `r round(sd_dur_s1, 2)`). The sample was composed of 100 female participants, 97 male, and 3 that declared otherwise. Mean age was `r round(mean_age_s1, 2)` (SD = `r round(sd_age_s1, 2)`).

## Materials

**Experimental Stimuli**

Participants were presented with introductory vignettes designed to prime their focus on a specific target politician (Donald Trump or Joe Biden). These paragraphs provided brief biographical details, including party affiliation and term dates, and prompted reflection on leadership and policy.

For the exploratory phase, a news vignette was used to describe the discovery of classified documents. The text was identical for both conditions, with only the politician’s name modified:

> "In late 2022 and early 2023, authorities discovered classified documents in \[Politician\]'s former office and personal residence... sparking an investigation into potential unauthorized retention of national defense information."

**Citizen Trust in Government Scale**

Trust was measured using an adaptation of the scale developed by @grimmelikhuijsenValidatingScaleCitizen2017. The scale consists of nine items measuring three dimensions:

-   **Perceived Competence:** (e.g., "\[Politician\] carries out his duty very well")

-   **Benevolence:** (e.g., "\[Politician\] is genuinely interested in the wellbeing of citizens")

-   **Integrity:** (e.g., "\[Politician\] approaches citizens in a sincere way")

Responses were recorded on a 5-point Likert scale ranging from 1 (*Strongly Disagree*) to 5 (*Strongly Agree*).

**Political Identification and Demographics**

Political affiliation was assessed through a multi-step branching sequence. Participants first identified their broad affiliation (Democrat, Republican, Independent, Other, or No preference).

-   **Partisans** were asked to rate the strength of their affiliation (*Strong* vs. *Not very Strong*).

-   **Non-partisans** were asked if they leaned toward either major party, with "Neither" provided as an option.

-   **Identity Importance:** A 5-point Likert scale measured how central party identification was to the participant’s self-concept (*Not at all important* to *Extremely important*).

## Procedure

The study was conducted online via the Qualtrics platform. After providing informed consent, participants were randomly assigned to either the **Donald Trump** or **Joe Biden** condition.

The experimental flow proceeded as follows:

1.  **Priming:** Participants read the introductory paragraph corresponding to their assigned politician.
2.  **Initial Assessment (Pre-test):** Participants completed the adapted Citizen Trust in Government scale regarding the target politician.
3.  **Exploratory Intervention:** Participants were presented with the news vignette regarding the classified documents scandal.
4.  **Second Assessment (Post-test):** Participants rated their trust in the politician a second time. During this step, they were shown their previous responses as a reference point to allow for intentional adjustment or consistency.
5.  **Classification:** Participants completed the political affiliation branching questions and the identity importance scale.
6.  **Demographics and Validation:** Participants provided demographic information (age, gender) and completed two attention check items.

Upon completion, all participants were fully debriefed. For the purpose of the primary analyses, the Joe Biden condition was treated as the reference category.

## Data Analysis

All participants passed both attention checks, and data cleaning was performed as pre-registered, that is, we excluded the fastest and slowest one percent of the sample, leading to a sample size of 197. We used linear regression to assess whether people's political affiliation predicts baseline perceptions of trust. The average score across all trust items acts as the dependent variable. As the independent variable, we took participants' political affiliation, which was informed by participants' *Prolific* profiles. For the purposes of these analyses, Democrat affiliation is the reference category. The baseline linear regression model assesses the impact of political affiliation on trust:

$$
\text{Trust}_{1i} = \beta_0 + \beta_1 \cdot \text{Political Affiliation}_i + \beta_2 \cdot \text{Politician}_i + \beta_3 \cdot \text{Political Affiliation}_i \cdot \text{Politician}_i + \epsilon_i
$$

We also used linear regression to evaluate whether the change in perceptions of trust is affected by participants' political affiliation. In this case, the dependent variable is the change in perceived trust, calculated as the difference between the second measurement of trust after being informed about a scandal and the first measurement (before). The regression is as follows:

\begin{align*}
\Delta \text{Trust}_{i} &= \text{Trust}_{2i} - \text{Trust}_{1i} \\
&= \beta_0 + \beta_1 \cdot \text{Political Affiliation}_i + \beta_2 \cdot \text{Politician}_i \\
&\quad + \beta_3 \cdot (\text{Political Affiliation}_i \cdot \text{Politician}_i) + \epsilon_i
\end{align*}

## Results

### Baseline Levels of Trust

```{r s1-violin-trust, fig.cap="Trust by Politician and Political Affiliation (Study 1)."}
ggplot(df1, aes(x = Politician, y = Trust1, fill = PoliticalAffiliation)) +
  geom_violin(alpha = 0.6, trim = FALSE, position = position_dodge(width = 0.9)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 21,
    size = 4,
    position = position_dodge(width = 0.9)
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    width = 0.2,
    position = position_dodge(width = 0.9)
  ) +
  labs(
    title = "Violin Plot of Trust by Politician\nand Political Affiliation",
    x = "Politician", y = "Trust", fill = "Political Affiliation"
  ) +
  scale_fill_manual(values = c("Democrat" = "blue", "Republican" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.5),
    panel.grid.minor.y = element_blank()
  )
```

```{r s1-violin-change, fig.cap="Trust Change by Politician and Political Affiliation after exposure to a negative news vignette (Study 1)."}
ggplot(df1, aes(x = Politician, y = trust_change, fill = PoliticalAffiliation)) +
  geom_violin(alpha = 0.6, trim = FALSE, adjust = 2, position = position_dodge(width = 0.9)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 21,
    size = 4,
    position = position_dodge(width = 0.9)
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    width = 0.2,
    position = position_dodge(width = 0.9)
  ) +
  labs(
    title = "Violin Plot of Trust Change\nby Politician and Political Affiliation",
    x = "Politician", y = "Trust Change", fill = "Political Affiliation"
  ) +
  scale_fill_manual(values = c("Democrat" = "blue", "Republican" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.5),
    panel.grid.minor.y = element_blank()
  )
```

```{r s1-models, results='hide'}
lm1 <- lm(trust1 ~ dummy_polaff + dummy_pol + dummy_pol*dummy_polaff, data = df1)
apa_lm1 <- apa_print(lm1)
lm2 <- lm(trust_change ~ dummy_polaff + dummy_pol + dummy_pol*dummy_polaff, data = df1)
apa_lm2 <- apa_print(lm2)
```

The linear regression results reveal a strong and statistically significant interaction between participants' political affiliation and the politician they evaluated in predicting baseline trust. The intercept indicates that Democrats evaluating Joe Biden reported relatively high trust (`r apa_lm1$estimate[1]`). Trust was significantly lower when the participant was a Republican (`r apa_lm1$estimate[2]`) or when the evaluated politician was Donald Trump (`r apa_lm1$estimate[3]`). Importantly, the positive interaction term (`r apa_lm1$estimate[4]`) shows that Republicans evaluating Trump also report high levels of trust for their own representative. This suggests that the alignment between a Republican participant and Trump results in a substantial boost in trust perceptions, highlighting a strong partisan congruence effect. Results are reported in *Table 1*.

### Change in Trust after Learning about Bad News

We now apply the same analysis structure as above, but with change in trust as the dependent variable. The intercept is highly significant (`r apa_lm2$estimate[1]`), which shows that on average, trust ratings reduced after exposure to the negative news. Furthermore, there is a non-significant effect for Trump vs. Biden (`r apa_lm2$estimate[3]`), such that when the bad news is about Trump, trust does not drop as much as it does when the bad news is about Biden. There are no significant differences in changes in Trust. Results are reported in *Table 1*. These regression coefficients imply that a Democrat who read about a Biden scandal shifted their Trust by `r round(lm2$coefficients[1], 2)`; a Republican who learned about the Biden scandal shifted by `r round(lm2$coefficients[1] + lm2$coefficients[2], 2)`; a Democrat who read about Trump shifted by `r round(lm2$coefficients[1] + lm2$coefficients[3], 2)`; and a Republican who read about a Trump scandal shifted by `r round(lm2$coefficients[1] + lm2$coefficients[2] + lm2$coefficients[3] + lm2$coefficients[4], 2)`. It is noteworthy that baseline levels of trust in Trump among Democrat participants were already markedly low, suggesting the possibility of a floor effect whereby the subsequent scandal had limited additional impact, as there was little room for trust to further decrease.

```{r s1-regression-table, results='asis'}
stargazer(lm1, lm2,
          type = 'latex',
          header = FALSE,
          dep.var.labels = c("Trust", "Change in Trust"),
          covariate.labels = c("Political Affiliation (Republican = 1)", "Dummy Politician (Trump = 1)", "Interaction"))
```

### Principal Component Analysis

```{r s1-pca, results='hide'}
df1_subset <- df1[, c('Q1_1', 'Q1_2', 'Q1_3', 'Q1_4', 'Q1_5', 'Q1_6', 'Q1_7', 'Q1_8', 'Q1_9')]
pca1dim <- principal(df1_subset, nfactors = 1, rotate = "varimax", scores = TRUE)
pca1dim
```

```{r s1-scree, fig.cap="Scree plot of the Principal Component Analysis for Study 1. The red dashed line represents the Kaiser criterion (eigenvalue = 1). The blue labels show the cumulative variance explained by components above the criterion."}
eigenvalues <- pca1dim$values
scree_data <- data.frame(
  Component = 1:length(eigenvalues),
  Eigenvalues = eigenvalues,
  CumulativeVariance = cumsum(eigenvalues) / sum(eigenvalues)
)

ggplot(scree_data, aes(x = Component, y = Eigenvalues)) +
  geom_line() +
  geom_point() +
  theme_classic() +
  labs(title = "Scree Plot of Principal Component Analysis",
       x = "Principal Component",
       y = "Eigenvalues") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_text(aes(label = ifelse(Eigenvalues > 1, round(CumulativeVariance, 2), "")),
            vjust = -1, size = 3, color = "blue")
```

In order to better understand how the different dimensions of Trust are represented in perceptions of these two politicians within a highly polarised context, a Principal Component Analysis (PCA) was conducted. This analysis allows us to test whether the Trust measure retains its hypothesized three-component structure across different contexts. The PCA results revealed a clear one-factor solution, suggesting that the items load onto a single underlying construct. The one-dimensional model explained 89% of the total variance, with a steep drop in the scree plot after the first component (see @fig-s1-scree), indicating a strong dominant factor. While the two-dimensional solution accounted for slightly more variance (92%), the gain was marginal without a clear interpretive advantage. The three-dimensional model increased the explained variance to 94%, but each additional component contributed progressively less.

This finding suggests that, in a highly polarised U.S. sample, people did not differentiate between competence, benevolence, and integrity when rating a known political figure. Instead, their answers to all items were highly correlated, reflecting a global evaluation: either they broadly trust the politician or they do not. In essence, trust functioned as a unidimensional attitude in this context. Given the nature of the study—in which Democrat and Republican participants evaluated either Joe Biden or Donald Trump—we can argue that partisan polarisation reduced the dimensionality of trust: participants who disliked the opposing party's leader rated them low on every aspect, whereas those who liked their in-group leader rated them high across the board. This provides initial empirical support for the hypothesis that polarisation reduces trust dimensionality as a result of lower deliberation (i.e., more System 1 thinking).

## Discussion

Study 1 established that, in a polarised U.S. context, political affiliation is a pivotal determinant of trust in politicians. People placed their trust in politicians who shared their political affiliation and discounted trust when they did not. Despite this partisan rigidity, respondents were still responsive to new negative information—trust decreased following exposure to a scandal vignette. Moreover, the PCA suggests that trust can be effectively represented as a unidimensional construct in this context, with a single overarching dimension capturing the majority of the variance. Study 1 established the phenomenon; in the studies that follow, we examine whether it generalises beyond the U.S.

# Study 2: Replicating the Trust Structure in Chile

## Methods

All methods were pre-registered before data collection in [aspredicted.org #257474](https://aspredicted.org/s4b638.pdf).

## Participants

```{r s2-load, include=FALSE}
d2_raw <- read_csv("study2.csv")

# Remove Qualtrics metadata rows (first row is header, second is question text)
d2 <- d2_raw %>%
  mutate(
    duration = as.numeric(`Duration (in seconds)`),
    Politician = Politician,
    izq_der = as_factor(`Políticamente hablando, ¿te identificas más con la Izquierda o la Derecha? - Selected Choice`),
    c1 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] es capaz.`,
    c2 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] es un/a experto/a.`,
    c3 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] cumple muy bien con su deber.`,
    b1 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - Si los ciudadanos necesitan ayuda, [Field-Politician] hará lo posible para ayudarles.`,
    b2 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] actúa en interés de los ciudadanos.`,
    b3 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] está genuinamente interesado/a en el bienestar de los ciudadanos.`,
    i1 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] tiene un acercamiento sincero con los ciudadanos.`,
    i2 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] es sincero/a.`,
    i3 = `Cuando se trata de [Field-Politician], ¿hasta qué punto estás de acuerdo o en desacuerdo con las siguientes afirmaciones? - [Field-Politician] es honesto/a.`,
    age = as.numeric(`¿Qué edad tienes?`),
    gender = `¿Con qué género te identificas? - Selected Choice`
  )

# Filter out extreme durations
d2 <- d2 %>%
  filter(between(duration,
                 quantile(duration, 0.01, na.rm = TRUE),
                 quantile(duration, 0.99, na.rm = TRUE)))

# Recode Likert scale
recode_likert_es <- function(x) {
  case_when(
    x == "Totalmente en desacuerdo" ~ 1,
    x == "Algo en desacuerdo" ~ 2,
    x == "Ni de acuerdo ni en desacuerdo" ~ 3,
    x == "Algo de acuerdo" ~ 4,
    x == "Totalmente de acuerdo" ~ 5,
    TRUE ~ NA_real_
  )
}

vars_trust <- c("c1", "c2", "c3", "b1", "b2", "b3", "i1", "i2", "i3")
d2 <- d2 %>%
  mutate(across(all_of(vars_trust), recode_likert_es))

d2 <- d2 %>%
  mutate(trust = rowMeans(select(., all_of(vars_trust)), na.rm = TRUE))

n_s2 <- nrow(d2)
mean_age_s2 <- mean(d2$age, na.rm = TRUE)
sd_age_s2 <- sd(d2$age, na.rm = TRUE)
mean_dur_s2 <- mean(d2$duration, na.rm = TRUE)
sd_dur_s2 <- sd(d2$duration, na.rm = TRUE)
```

We recruited `r n_s2` participants from a Cint panel between November 12 and 16, 2025, coinciding with the Chilean presidential elections. We recruited participants who identified as either left-leaning or right-leaning. The actual duration was `r round(mean_dur_s2, 2)` seconds (SD = `r round(sd_dur_s2, 2)`). Mean age was `r round(mean_age_s2, 2)` (SD = `r round(sd_age_s2, 2)`).

## Materials

**Experimental Stimuli**

Participants were randomly assigned to evaluate one of two prominent figures in the 2025 Chilean presidential election: **José Antonio Kast** (Republican Party, Right) or **Jeannette Jara** (Communist Party, Left). Unlike Study 1, this study focused on baseline trust levels without the introduction of an experimental scandal vignette.

**Localized Citizen Trust Scale**

Trust was measured using a localized Spanish-language adaptation of the scale by @grimmelikhuijsenValidatingScaleCitizen2017. The nine-item instrument was modified to fit the Chilean political context while maintaining the three hypothesized dimensions.

## Procedure

The study was implemented via Qualtrics following a cross-sectional design focused on the dimensionality of trust.

The experimental flow was as follows:

1.  **Randomization:** Participants were assigned to either the **Kast** or **Jara** condition.
2.  **Trust Assessment:** Participants completed the nine-item localized trust scale regarding their assigned politician. No pre/post manipulation was used; only baseline trust was recorded.
3.  **Political Identification:** Participants reported their political affiliation and identity strength. For the subsequent analyses:
    -   **Jeannette Jara** was established as the reference category for the target politician.
    -   **Left-identifying participants** were established as the reference group for partisanship.
4.  **Demographics and Validation:** Participants provided demographic data and completed attention checks before being debriefed.

## Data Analysis

Data cleaning followed pre-registered criteria: we excluded the fastest and slowest one percent of the sample based on response duration. We used Principal Component Analysis (PCA) to examine the dimensional structure of trust perceptions in this sample. We also analyzed mean trust differences between in-group and out-group conditions to verify the presence of polarisation effects.

## Results

### Violin Plot of Trust

```{r s2-violin, fig.cap="Trust by Politician and Political Orientation in Chile (Study 2)."}
ggplot(d2, aes(x = Politician, y = trust, fill = izq_der)) +
  geom_violin(alpha = 0.5, position = position_dodge(width = 0.9), trim = TRUE) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 21,
    size = 3,
    color = "black",
    position = position_dodge(width = 0.9)
  ) +
  geom_boxplot(
    width = 0.2,
    outlier.shape = NA,
    alpha = 0.5,
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  scale_fill_manual(values = c("Izquierda" = "red", "Derecha" = "blue")) +
  labs(
    title = "Trust by Politician and Political Orientation",
    x = "Politician",
    y = "Trust (1-5)",
    fill = "Political Orientation"
  ) +
  theme_classic()
```

### Principal Component Analysis

```{r s2-pca, results='hide'}
d2_pca <- d2 %>%
  select(all_of(vars_trust)) %>%
  na.omit()

pca_s2 <- principal(d2_pca, nfactors = 1, rotate = "varimax", scores = TRUE)
pca_s2
```

```{r s2-scree, fig.cap="Scree plot of the Principal Component Analysis for Study 2 (Chile). The red dashed line represents the Kaiser criterion (eigenvalue = 1). The blue labels show the cumulative variance explained by components above the criterion."}
eigenvalues_s2 <- pca_s2$values
scree_data_s2 <- data.frame(
  Component = 1:length(eigenvalues_s2),
  Eigenvalues = eigenvalues_s2,
  CumulativeVariance = cumsum(eigenvalues_s2) / sum(eigenvalues_s2)
)

ggplot(scree_data_s2, aes(x = Component, y = Eigenvalues)) +
  geom_line() +
  geom_point() +
  theme_classic() +
  labs(title = "Scree Plot of Principal Component Analysis",
       x = "Principal Component",
       y = "Eigenvalues") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_text(aes(label = ifelse(Eigenvalues > 1, round(CumulativeVariance, 2), "")),
            vjust = -1, size = 3, color = "blue")
```

We conducted a Principal Component Analysis (PCA) on the nine-item trust scale to evaluate the dimensionality of trust in the Chilean context. Similar to the findings in the U.S., the analysis yielded a clear one-factor solution. This single component explained `r round(pca_s2[["Vaccounted"]][2], 2)*100`% of the total variance, with an eigenvalue of `r round(pca_s2[["Vaccounted"]][1], 2)`.

Standardized loadings for all items were exceptionally high, ranging from `r round(pca_s2[["loadings"]][2], 2)` to `r round(pca_s2[["loadings"]][4], 2)`, with a mean item complexity of `r round(mean(pca_s2[["complexity"]]), 2)`. The scree plot shows a sharp decline after the first component, which again supports that one dimension is adequate for this construct. Statistical fit was confirmed by a non-significant empirical chi-square $(\chi^2=$ `r round(pca_s2[["chi"]], 2)`$, p=$ `r round(pca_s2[["EPVAL"]], 2)`$)$ and a Root Mean Square of the Residuals (RMSR) of `r round(pca_s2[["rms"]], 2)`.

## Discussion

The replication of a unidimensional trust structure in Chile supports the hypothesis that high societal tensions and polarisation lead to a collapse in trust dimensionality. Despite Chile's different party system and lower formal party identification, the results suggest that when political identities become salient, they produce cognitive effects analogous to those in the U.S. These findings imply that simplified trust judgments are common features of polarised democracies, regardless of specific institutional histories.

# Study 3: Testing in Denmark's Low-Polarisation Environment

## Methods

All methods were pre-registered in [aspredicted.org #270,175](https://aspredicted.org/ju4737.pdf).

## Participants

```{r s3-load, include=FALSE}
d3_raw <- read_csv("study3.csv")

# Remove Qualtrics metadata rows
d3 <- d3_raw %>% slice(3:n())

d3 <- d3 %>%
  mutate(
    duration = as.numeric(`Duration (in seconds)`),
    block_aff = as_factor(affiliation),
    politician_check = attention_check,
    c1 = trust_1,
    c2 = trust_2,
    c3 = trust_3,
    b1 = trust_4,
    b2 = trust_5,
    b3 = trust_6,
    i1 = trust_7,
    i2 = trust_8,
    i3 = trust_9,
    age = as.numeric(age)
  )

# Filter extreme durations
d3 <- d3 %>%
  filter(between(duration,
                 quantile(duration, 0.01, na.rm = TRUE),
                 quantile(duration, 0.99, na.rm = TRUE)))

# Filter for only "Rød blok" or "Blå blok"
d3 <- d3 %>%
  filter(block_aff %in% c("Rød blok", "Blå blok"))

# Filter attention check
d3 <- d3 %>%
  filter(politician_check == Politician)

# Recode Likert scale (Danish)
recode_likert_dk <- function(x) {
  case_when(
    x == "Meget uenig" ~ 1,
    x == "Delvist Uenig" ~ 2,
    x == "Hverken enig eller uenig" ~ 3,
    x == "Delvist enig" ~ 4,
    x == "Meget enig" ~ 5,
    TRUE ~ NA_real_
  )
}

vars_trust <- c("c1", "c2", "c3", "b1", "b2", "b3", "i1", "i2", "i3")
d3 <- d3 %>%
  mutate(across(all_of(vars_trust), recode_likert_dk))

d3 <- d3 %>%
  mutate(trust = rowMeans(select(., all_of(vars_trust)), na.rm = TRUE))

n_s3 <- nrow(d3)
mean_age_s3 <- mean(d3$age, na.rm = TRUE)
sd_age_s3 <- sd(d3$age, na.rm = TRUE)
mean_dur_s3 <- mean(d3$duration, na.rm = TRUE)
sd_dur_s3 <- sd(d3$duration, na.rm = TRUE)
```

We recruited `r n_s3` participants from *Cint*. We filtered for participants who identified with either Rød blok (Red bloc) or Blå blok (Blue bloc). The actual duration was `r round(mean_dur_s3, 2)` seconds (SD = `r round(sd_dur_s3, 2)`). Mean age was `r round(mean_age_s3, 2)` (SD = `r round(sd_age_s3, 2)`).

## Materials

**Experimental Stimuli**

Participants were randomly assigned to evaluate one of two prominent figures in Danish politics: **Mette Frederiksen** (Social Democrats, Center-Left) or **Troels Lund Poulsen** (Venstre, Center-Right). Similar to Studies 1 and 2, participants read a brief introductory paragraph providing the politician's party affiliation and current role to standardize the context before assessment.

**Danish Citizen Trust Scale**

Trust was measured using a Danish-translated version of the nine-item scale by @grimmelikhuijsenValidatingScaleCitizen2017. The instrument assessed three dimensions: perceived competence, benevolence, and integrity. Participants indicated their agreement on a 5-point Likert scale from *Meget uenig* (Strongly Disagree) to *Meget enig* (Strongly Agree).

## Procedure

The study was hosted on the Qualtrics platform. After providing informed consent, participants were randomly assigned to either the **Frederiksen** or **Poulsen** condition.

The experimental procedure followed this sequence:

1.  **Stimuli:** Participants were presented with the introductory biographical paragraph for their assigned politician.
2.  **Trust Assessment:** Immediately following the prime, participants completed the nine-item Danish trust scale.
3.  **Validation:** Before proceeding to the final section, participants completed two attention check items to ensure data quality.
4.  **Demographics and Political Identity:** Participants provided demographic information (e.g., age, gender) and reported their political affiliation, strength of that affiliation, and importance.

Upon completion, all participants were fully debriefed.

## Data Analysis

Data cleaning followed the same criteria as Studies 1 and 2: we excluded the fastest and slowest one percent of the sample based on response duration and retained only participants who passed the attention check. We conducted a Principal Component Analysis (PCA) to test the dimensional structure of trust perceptions in the Danish sample.

## Results

### Violin Plot of Trust

```{r s3-violin, fig.cap="Trust by Politician and Political Bloc in Denmark (Study 3)."}
ggplot(d3, aes(x = Politician, y = trust, fill = block_aff)) +
  geom_violin(alpha = 0.5, position = position_dodge(width = 0.9), trim = TRUE) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 21,
    size = 3,
    color = "black",
    position = position_dodge(width = 0.9)
  ) +
  geom_boxplot(
    width = 0.2,
    outlier.shape = NA,
    alpha = 0.5,
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  scale_fill_manual(values = c("Rød blok" = "red", "Blå blok" = "blue")) +
  labs(
    title = "Trust by Politician and Political Affiliation",
    x = "Politician",
    y = "Trust (1-5)",
    fill = "Political Affiliation"
  ) +
  theme_classic()
```

### Principal Component Analysis

```{r s3-pca, results='hide'}
d3_pca <- d3 %>%
  select(all_of(vars_trust)) %>%
  na.omit()

pca_s3 <- principal(d3_pca, nfactors = 1, rotate = "varimax", scores = TRUE)
pca_s3
```

```{r s3-scree, fig.cap="Scree plot of the Principal Component Analysis for Study 3 (Denmark). The red dashed line represents the Kaiser criterion (eigenvalue = 1). The blue labels show the cumulative variance explained by components above the criterion."}
eigenvalues_s3 <- pca_s3$values
scree_data_s3 <- data.frame(
  Component = 1:length(eigenvalues_s3),
  Eigenvalues = eigenvalues_s3,
  CumulativeVariance = cumsum(eigenvalues_s3) / sum(eigenvalues_s3)
)

ggplot(scree_data_s3, aes(x = Component, y = Eigenvalues)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(title = "Scree Plot of Principal Component Analysis",
       x = "Principal Component",
       y = "Eigenvalues") +
  theme_classic() +
  geom_text(aes(label = ifelse(Eigenvalues > 1, round(CumulativeVariance, 2), "")),
            vjust = -1, size = 3, color = "blue")
```

The PCA conducted on the Danish sample produced an unexpected unidimensional result, contrary to the initial hypothesis of a three-dimensional structure. The single factor explained 78% of the total variance, with a dominant eigenvalue of 7.01.

Loadings for competence, benevolence, and integrity items remained high (.83 to .90), and the mean item complexity was 1.0. The model fit was statistically adequate ($\chi^2=25.08$, $p<.57$; RMSR = 0.03).

## Discussion

The emergence of a single factor in Denmark—a low-polarisation environment—suggests that the collapse of trust dimensionality may be more fundamental to political judgment than previously theorised. While polarisation certainly exacerbates this effect in the U.S. and Chile, the Danish results indicate that even in high-trust, consensus-oriented cultures, citizens evaluate specific political leaders through a global lens rather than differentiating between competence and character. This implies that trust in individual politicians may naturally function as a visceral, System 1 heuristic, regardless of the broader political climate.

# General Discussion

Across three countries—the United States, Chile, and Denmark—we consistently found that trust in politicians functions as a unidimensional construct. This consistency represents the central and unexpected finding of this research programme.

## The Pervasiveness of Unidimensional Trust

The most striking result is the uniformity of the one-factor solution across contexts that differ substantially in political culture, party system, and polarisation levels. Our original hypothesis anticipated that polarisation would be the key driver of trust dimensionality reduction—that highly polarised settings (U.S. and Chile) would produce collapsed, heuristic-driven trust judgments, while the low-polarisation, high-trust environment of Denmark would permit more deliberate, multidimensional assessments. The Danish data did not bear this out. Even among participants who identify with distinct political blocs and who evaluate politicians on opposite sides of the ideological spectrum, trust collapsed onto a single evaluative dimension just as it did elsewhere.

One possible alternative interpretation is that the politicians evaluated in Denmark are perceived as comparably competent, benevolent, and honest by those who support them, such that the scale fails to produce variance along separate dimensions. Although this cannot be ruled out, we believe the cross-country consistency of results points to something more fundamental. It may be the case that the @grimmelikhuijsenValidatingScaleCitizen2017 scale—originally validated for citizen trust in government *organisations* rather than individual political figures—does not adequately capture multidimensional trust when the object of judgment is a salient individual rather than an abstract institution. The interpersonal nature of politician evaluation may reliably activate a global affective response that overrides analytical differentiation of competence, benevolence, and integrity, regardless of the broader political context.

## Theoretical Implications

These results challenge the universal applicability of multidimensional trust models [e.g., @mayerIntegrativeModelOrganizational1995; @grimmelikhuijsenValidatingScaleCitizen2017] when applied to prominent political figures. While citizens may be capable of distinguishing trust dimensions when evaluating abstract institutions, they appear to collapse them when evaluating named individuals. This is consistent with a Halo Effect interpretation [@thorndike1920]: a general positive or negative sentiment toward a politician dictates responses across all sub-dimensions of trust. From a dual-process perspective, our findings suggest that even when System 2 engagement is theoretically more likely—as in a low-polarisation, high-trust society—the act of judging a familiar political figure may reliably recruit System 1 heuristics. The salience of individual politicians as social and partisan objects may be sufficient to short-circuit deliberate, dimensional evaluation.

These results also carry practical implications. If citizens consistently collapse trust into a single dimension, then policy efforts aimed at building trust through "demonstrated competence" alone are unlikely to be effective if a baseline affective deficit exists toward a leader. Communication strategies must first address the global affective orientation of the public toward a political figure before specific attributes such as integrity or expertise can register as distinct positive signals.

## Limitations and Future Directions

Several limitations should be noted. First, the studies relied on relatively short, single-session survey designs. Longitudinal approaches might reveal whether prolonged exposure to political figures or shifting political contexts can restore dimensionality in trust judgments. Second, the reliance on the @grimmelikhuijsenValidatingScaleCitizen2017 scale, while enabling direct comparisons across studies, may limit our ability to detect multidimensionality if the instrument itself is not sensitive enough when applied to individual political figures. Future research should explore alternative measurement strategies—including implicit measures or process-tracing methods—to better capture the latent structure of political trust. Third, the selection of specific politicians in each country may have influenced results; replication with different target figures would strengthen the generalisability of these findings.

Future research should also examine the causal mechanisms linking political context to trust structure, consider whether the unidimensionality of trust varies by political involvement or cognitive style, and explore whether interventions designed to promote deliberate evaluation can restore dimensional differentiation even in polarised contexts.

## Conclusion

Our research demonstrates that political trust, as measured by a validated nine-item scale, is effectively a unidimensional attitude across contemporary democratic contexts that vary widely in polarisation and political culture. In highly polarised environments like the U.S. and Chile, this reduction is likely reinforced by affective animosity and partisan identity. However, the persistence of this structure in Denmark suggests that unidimensionality may be a baseline feature of how citizens process information about individual political leaders, rather than a specific product of polarisation. Understanding this structural simplicity is a prerequisite for designing communication strategies and trust-building initiatives that can resonate across the ideological divides of modern democratic societies.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
